---
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 1 Load Libraries

```{r}
library(tidyverse)
library(readxl)
```

# 2 Load and Clean Data

## 2.1 Long-term Yield

```{r}
# Read long-term yield, skip header rows
lt_raw <- read_csv("cranberry-data/data_longterm/LTYielddata2024.csv",
                   skip = 3, show_col_types = FALSE,
                   col_names = c("cultivar", "heat_treatment", "plot",
                                 "rotten_count", "rotten_weight",
                                 "healthy_count", "healthy_weight",
                                 "total_count", "total_weight"))

lt <- lt_raw %>%
  filter(!is.na(cultivar), cultivar != "Cultivar") %>%
  mutate(across(c(plot, rotten_count:total_weight), as.numeric),
         heat_treatment = str_trim(heat_treatment),
         pct_rotten_wt = rotten_weight / total_weight * 100,
         pct_rotten_ct = rotten_count / total_count * 100)

cat("Long-term yield:", nrow(lt), "rows\n")
print(lt)
```

## 2.2 Acute Yield

```{r}
# Read acute yield, skip header rows
acute_raw <- read_excel("cranberry-data/data_acute/Acute HS-Yield_RawData 2024.xlsx",
                        sheet = "Acute Heat stress", skip = 4)

acute <- acute_raw %>%
  select(
    cultivar = Cultivar,
    trt_string = Treatment,
    bio_rep = `# OTC/ plot`,
    rotten_count = `Rotten fruit #`,
    rotten_weight = `Rotten fruitWeight (g)`,
    healthy_count = `Non-rotten fruit#`,
    healthy_weight = `Non-rotten fruitWeight (g)`,
    total_count = `Total #`,
    total_weight = `Total Weight (g)`
  ) %>%
  filter(!is.na(trt_string), !str_detect(trt_string, "A0")) %>%
  mutate(
    # "C" alone = treatment C (OTC); "AC","BC","CC","DC" = controls
    is_control = nchar(trt_string) >= 2 & str_detect(trt_string, "C$"),
    treatment = str_extract(trt_string, "^[A-D]"),
    plot_type = if_else(is_control, "Control", "OTC"),
    bio_rep = as.integer(bio_rep),
    pct_rotten_wt = rotten_weight / total_weight * 100,
    pct_rotten_ct = rotten_count / total_count * 100
  )

cat("Acute yield (excluding A0):", nrow(acute), "rows\n")
print(acute %>% count(cultivar, treatment, plot_type))
```

# 3 Exploratory Data Analysis

## 3.1 Long-term: OTC vs Control

```{r lt-eda, fig.height=4}
# Healthy weight by treatment and cultivar
ggplot(lt, aes(x = heat_treatment, y = healthy_weight, fill = heat_treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Long-term: Healthy Fruit Weight",
       x = NULL, y = "Weight (g)") +
  theme_minimal() + theme(legend.position = "none")
```

For Stevens, OTC and Control healthy weights are nearly identical (~684g vs ~681g).
For MQ, OTC yield is lower than Control (835g vs 961g), suggesting MQ may be
more sensitive to long-term heating, though variability is high with only 4 reps.

```{r lt-rotten, fig.height=4}
# Rotten percentage
ggplot(lt, aes(x = heat_treatment, y = pct_rotten_wt, fill = heat_treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Long-term: % Rotten by Weight",
       x = NULL, y = "% Rotten") +
  theme_minimal() + theme(legend.position = "none")
```

Interestingly, OTC plots have lower rotten rates in both cultivars. MQ has
higher rotten rates overall (2-8%) compared to Stevens (0-4%). The OTC
structure may provide some protection against rot-causing conditions.

## 3.2 Acute: Treatment Effect

```{r acute-hw, fig.height=5}
# Healthy weight by treatment, split by OTC/Control and cultivar
ggplot(acute, aes(x = treatment, y = healthy_weight, fill = plot_type)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(aes(color = plot_type), position = position_jitterdodge(jitter.width = 0.1),
              size = 1.5) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  scale_color_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Acute: Healthy Fruit Weight by Treatment",
       x = "Treatment (A=4, B=3, C=2, D=1 pulse)", y = "Weight (g)") +
  theme_minimal() + theme(legend.position = "bottom")
```

OTC (orange) consistently produces lower healthy weight than Control (blue)
across nearly all treatments and cultivars. The effect is visually clear.
However, there is no obvious dose-response pattern -- A (4 pulses) does not
yield notably less than D (1 pulse). St-B is the only case where OTC > Control.

```{r acute-rotten, fig.height=5}
# Rotten percentage
ggplot(acute, aes(x = treatment, y = pct_rotten_wt, fill = plot_type)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(aes(color = plot_type), position = position_jitterdodge(jitter.width = 0.1),
              size = 1.5) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  scale_color_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Acute: % Rotten by Weight",
       x = "Treatment (A=4, B=3, C=2, D=1 pulse)", y = "% Rotten") +
  theme_minimal() + theme(legend.position = "bottom")
```

OTC plots show dramatically higher rotten percentages than Control in every
treatment group. Control rotten rates cluster around 1-5%, while OTC rates
range from 5-37%. Treatment D (1 pulse) shows the highest variability and
some extreme outliers (>25% rotten), suggesting single late-season pulses
may be particularly damaging.

## 3.3 Acute: Paired OTC-Control Difference

```{r acute-diff, fig.height=5}
# Compute OTC - Control difference per treatment/cultivar/bio_rep pair
acute_paired <- acute %>%
  select(cultivar, treatment, bio_rep, plot_type, healthy_weight, pct_rotten_wt) %>%
  pivot_wider(names_from = plot_type,
              values_from = c(healthy_weight, pct_rotten_wt)) %>%
  mutate(diff_hw = healthy_weight_OTC - healthy_weight_Control,
         diff_rotten = pct_rotten_wt_OTC - pct_rotten_wt_Control)

ggplot(acute_paired, aes(x = treatment, y = diff_hw, color = cultivar)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_jitter(width = 0.15, size = 3, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4,
               position = position_dodge(0.4)) +
  labs(title = "Acute: OTC - Control Healthy Weight Difference",
       subtitle = "Negative = heat reduced yield",
       x = "Treatment", y = "Difference (g)") +
  theme_minimal() + theme(legend.position = "bottom")
```

Nearly all paired differences are negative (below the dashed zero line),
confirming that OTC heating reduces yield. The mean reduction is about 155g.
The crossbars show group means -- all treatments show negative means except
St-B. There is no clear trend with pulse frequency.

# 4 Long-term Yield Modeling

## 4.1 Paired Design

Long-term experiment has 8 set pairs (OTC + Control). We use the set pairing
structure: plots 1-4 are OTC paired with plots 5-8 as Control (Stevens),
and plots 9-12 OTC paired with 13-16 Control (MQ).

```{r lt-pair}
# Assign set pair ID based on plot number
lt <- lt %>%
  mutate(set_pair = case_when(
    plot %in% c(1, 5)   ~ 1,
    plot %in% c(2, 6)   ~ 2,
    plot %in% c(3, 7)   ~ 3,
    plot %in% c(4, 8)   ~ 4,
    plot %in% c(9, 13)  ~ 5,
    plot %in% c(10, 14) ~ 6,
    plot %in% c(11, 15) ~ 7,
    plot %in% c(12, 16) ~ 8
  ))

# Compute paired differences
lt_wide <- lt %>%
  select(cultivar, set_pair, heat_treatment, healthy_weight, pct_rotten_wt) %>%
  pivot_wider(names_from = heat_treatment,
              values_from = c(healthy_weight, pct_rotten_wt)) %>%
  mutate(diff_hw = healthy_weight_OTC - healthy_weight_Control,
         diff_rotten = pct_rotten_wt_OTC - pct_rotten_wt_Control)

print(lt_wide %>% select(cultivar, set_pair, diff_hw, diff_rotten))
```

Paired differences vary widely: some pairs show OTC yielding more (+190g for
St pair 1), others show large reductions (-363g for MQ pair 8). This high
within-cultivar variability limits detection of a consistent OTC effect.

## 4.2 Paired t-test

```{r lt-ttest}
# Healthy weight: paired t-test (pooling both cultivars)
cat("=== Paired t-test: Healthy Weight (OTC - Control) ===\n")
t_hw <- t.test(lt_wide$diff_hw, mu = 0)
print(t_hw)

cat("\n=== Paired t-test: % Rotten (OTC - Control) ===\n")
t_rot <- t.test(lt_wide$diff_rotten, mu = 0)
print(t_rot)

# By cultivar
cat("\n=== By Cultivar: Healthy Weight ===\n")
for (cv in c("St", "MQ")) {
  sub <- lt_wide %>% filter(cultivar == cv)
  tt <- t.test(sub$diff_hw, mu = 0)
  cat(cv, ": mean diff =", round(mean(sub$diff_hw), 1),
      "g, p =", round(tt$p.value, 4), "\n")
}

cat("\n=== By Cultivar: % Rotten ===\n")
for (cv in c("St", "MQ")) {
  sub <- lt_wide %>% filter(cultivar == cv)
  tt <- t.test(sub$diff_rotten, mu = 0)
  cat(cv, ": mean diff =", round(mean(sub$diff_rotten), 1),
      "pp, p =", round(tt$p.value, 4), "\n")
}
```

None of the paired t-tests are significant (all p > 0.13). The overall mean
difference is -62g (OTC lower) but p=0.34. With only 4 pairs per cultivar,
statistical power is very limited. The direction suggests OTC may reduce MQ
yield (-126g) but has essentially no effect on Stevens (+3g).

## 4.3 Two-way ANOVA

```{r lt-anova}
# Model: healthy_weight ~ heat_treatment * cultivar
m_lt_hw <- lm(healthy_weight ~ heat_treatment * cultivar, data = lt)
cat("=== Long-term: Healthy Weight Model ===\n")
print(summary(m_lt_hw))

m_lt_rot <- lm(pct_rotten_wt ~ heat_treatment * cultivar, data = lt)
cat("\n=== Long-term: % Rotten Model ===\n")
print(summary(m_lt_rot))
```

The ANOVA confirms: cultivar is the only significant factor (p=0.002) --
Stevens yields ~280g less than MQ. The OTC effect is not significant
(p=0.11) but the coefficient (-126g for MQ) is practically meaningful.
The interaction is not significant (p=0.24), meaning both cultivars respond
similarly in direction if not magnitude. For rotten %, OTC shows a marginally
significant reduction of 2.5pp (p=0.064) in MQ.

# 5 Acute Yield Modeling

## 5.1 Model: Healthy Weight

```{r acute-model-hw}
# Main effects + interaction
m_acute_hw <- lm(healthy_weight ~ treatment * cultivar + plot_type,
                 data = acute)
cat("=== Acute: Healthy Weight ===\n")
print(summary(m_acute_hw))
print(anova(m_acute_hw))
```

The ANOVA table shows that only plot_type (OTC vs Control) is significant
(p=0.001). Treatment frequency and cultivar are not significant, nor is the
treatment-by-cultivar interaction (p=0.73).

## 5.2 Model: % Rotten

```{r acute-model-rot}
m_acute_rot <- lm(pct_rotten_wt ~ treatment * cultivar + plot_type,
                  data = acute)
cat("=== Acute: % Rotten ===\n")
print(summary(m_acute_rot))
print(anova(m_acute_rot))
```

Same pattern: plot_type is dominant (p < 0.001). Treatment and cultivar
effects are not significant, and the interaction is weak (p=0.55).

## 5.3 Simplified Models (drop non-significant interaction)

```{r acute-simplified}
# Check if interaction is needed
cat("=== Interaction test: Healthy Weight ===\n")
m_hw_no_int <- lm(healthy_weight ~ treatment + cultivar + plot_type,
                  data = acute)
print(anova(m_hw_no_int, m_acute_hw))

cat("\n=== Interaction test: % Rotten ===\n")
m_rot_no_int <- lm(pct_rotten_wt ~ treatment + cultivar + plot_type,
                   data = acute)
print(anova(m_rot_no_int, m_acute_rot))
```

The F-tests confirm the interaction terms are not needed (p=0.73 and p=0.55).
We proceed with the additive models.

```{r acute-final}
# Print the simpler model summaries
cat("=== Final Model: Healthy Weight ===\n")
print(summary(m_hw_no_int))

cat("\n=== Final Model: % Rotten ===\n")
print(summary(m_rot_no_int))
```

**Healthy weight model**: OTC reduces yield by 155g (p=0.001). Treatment
frequency has no significant effect -- B (+14g), C (+17g), D (+72g) relative
to A are all non-significant (p > 0.2). Cultivar is marginally non-significant
(St -69g, p=0.12). R-squared = 0.29.

**Rotten % model**: OTC increases rotten rate by 9.0 percentage points
(p < 0.001). Treatment D shows a trend toward higher rotten rate (+3.9pp,
p=0.15) but is not significant. Stevens has lower rotten rate (-3.1pp,
p=0.10). R-squared = 0.41.

# 6 Model Diagnostics

## 6.1 Long-term: Healthy Weight

```{r diag-lt, fig.height=7}
par(mfrow = c(2, 2))
plot(m_lt_hw, which = 1:4)
```

Residuals vs Fitted shows no obvious pattern. QQ plot looks acceptable for
n=16. No points exceed Cook's distance threshold. The diagnostics are
adequate given the small sample size.

## 6.2 Acute: Healthy Weight

```{r diag-acute-hw, fig.height=7}
par(mfrow = c(2, 2))
plot(m_hw_no_int, which = 1:4)
```

Residuals are approximately normal (QQ plot follows the line well). Residuals
vs Fitted shows relatively constant variance. A few points (rows 22, 24, 37)
have moderate leverage but Cook's distances remain below the threshold. This
model satisfies linear regression assumptions reasonably well.

## 6.3 Acute: % Rotten

```{r diag-acute-rot, fig.height=7}
par(mfrow = c(2, 2))
plot(m_rot_no_int, which = 1:4)
```

This model shows problems: the QQ plot reveals heavy right skew in residuals,
and the Scale-Location plot shows increasing variance. This is expected
because rotten percentages are bounded and right-skewed. Several points
(rows 22, 37, 46) are influential. A transformation is needed.

## 6.4 Influence Check

```{r influence}
# Check high Cook's distance points in acute rotten model
cooks_rot <- cooks.distance(m_rot_no_int)
influential <- which(cooks_rot > 4 / nrow(acute))
cat("Influential points (Cook's D > 4/n):\n")
print(acute[influential, c("cultivar", "trt_string", "bio_rep",
                            "healthy_weight", "pct_rotten_wt")])
```

Three observations have unusually high rotten rates (27-37%) compared to
typical values of 1-10%. These are real data points (not errors) -- they
represent plots where heat treatment caused severe fruit damage. We retain
them but note they drive much of the model's behavior.

## 6.5 Fix: Log-transform % Rotten

The rotten percentage residuals are non-normal (Shapiro p < 0.001, right-skewed).
We apply a log(y+1) transformation.

```{r fix-rotten}
acute <- acute %>% mutate(log_rotten = log(pct_rotten_wt + 1))

m_rot_log <- lm(log_rotten ~ treatment + cultivar + plot_type, data = acute)
cat("=== Log-transformed % Rotten Model ===\n")
print(summary(m_rot_log))

cat("\nShapiro-Wilk on log model residuals:\n")
print(shapiro.test(residuals(m_rot_log)))
```

The log-transformed model shows improved fit (R-squared = 0.62 vs 0.41).
OTC effect remains highly significant (p < 0.001), and now cultivar also
becomes significant (St has lower rotten rate, p=0.0007). Treatment D shows
a marginal trend (p=0.053). Shapiro-Wilk p=0.012 -- improved from 0.00004
but still somewhat non-normal due to the extreme observations.

```{r diag-rot-log, fig.height=7}
par(mfrow = c(2, 2))
plot(m_rot_log, which = 1:4)
```

Diagnostics are substantially improved after log transformation. The QQ plot
is closer to the reference line. Residual variance is more homogeneous.
A few mild outliers remain but Cook's distances are well-controlled.

# 7 Summary of Results

## 7.1 Long-term Experiment

```{r summary-lt, echo=FALSE}
cat("Long-term Heat Experiment (n=16, 4 OTC + 4 Control per cultivar)\n")
cat("Model: healthy_weight ~ heat_treatment * cultivar\n\n")

cat("Key findings:\n")
cat("- Cultivar effect is significant: St produces ~280g less",
    "than MQ (p=0.002)\n")
cat("- OTC heat treatment: MQ yield reduced by ~126g (p=0.11,",
    "not significant)\n")
cat("- OTC heat treatment: St yield essentially unchanged",
    "(+3g, p=0.97)\n")
cat("- OTC reduced rotten % in MQ by ~2.5pp (p=0.064,",
    "marginally significant)\n")
cat("- Limited sample size (n=4 per group) restricts",
    "statistical power\n")
```

## 7.2 Acute Experiment

```{r summary-acute, echo=FALSE}
cat("Acute Heat Experiment (n=48, 3 bio reps per group)\n")
cat("Model: healthy_weight ~ treatment + cultivar + plot_type\n\n")

cat("Key findings:\n")
cat("1. OTC heating significantly REDUCED healthy fruit weight",
    "by 155g (p=0.001)\n")
cat("   Baseline (MQ, Treatment A, Control): ~1011g\n")
cat("   OTC effect: -155g (~15% reduction)\n\n")

cat("2. OTC heating significantly INCREASED rotten fruit % by",
    "9.0pp (p<0.001)\n")
cat("   Baseline rotten rate: ~2.8%\n")
cat("   OTC effect: +9.0pp (rotten rate ~11.8% in OTC)\n\n")

cat("3. Treatment frequency (A=4, B=3, C=2, D=1 pulses) did",
    "NOT significantly affect yield\n")
cat("   No dose-response relationship detected (all p>0.2)\n\n")

cat("4. Cultivar differences were not statistically significant\n")
cat("   MQ tended to have higher yield (+69g, p=0.12)\n")
cat("   St tended to have lower rotten rate (-3.1pp, p=0.10)\n\n")

cat("Limitations:\n")
cat("- Small sample size (n=3 per treatment-cultivar-plot_type)\n")
cat("- % Rotten model has non-normal residuals even after",
    "log transform\n")
cat("- 3 influential observations with very high rotten rates",
    "(27-37%)\n")
```
