---
output:
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
```

# 1 Load Libraries

```{r}
library(tidyverse)
library(readxl)
library(coin
```

# 2 Load and Clean Data

## 2.1 Long-term Yield

```{r}
# Read long-term yield, skip header rows
lt_raw <- read_csv("cranberry-data/data_longterm/LTYielddata2024.csv",
                   skip = 3, show_col_types = FALSE,
                   col_names = c("cultivar", "heat_treatment", "plot",
                                 "rotten_count", "rotten_weight",
                                 "healthy_count", "healthy_weight",
                                 "total_count", "total_weight"))

lt <- lt_raw %>%
  filter(!is.na(cultivar), cultivar != "Cultivar") %>%
  mutate(across(c(plot, rotten_count:total_weight), as.numeric),
         heat_treatment = str_trim(heat_treatment),
         pct_rotten_wt = rotten_weight / total_weight * 100,
         pct_rotten_ct = rotten_count / total_count * 100)

cat("Long-term yield:", nrow(lt), "rows\n")
print(lt)
```

## 2.2 Acute Yield

```{r}
# Read acute yield, skip header rows
acute_raw <- read_excel("cranberry-data/data_acute/Acute HS-Yield_RawData 2024.xlsx",
                        sheet = "Acute Heat stress", skip = 4)

acute <- acute_raw %>%
  select(
    cultivar = Cultivar,
    trt_string = Treatment,
    bio_rep = `# OTC/ plot`,
    rotten_count = `Rotten fruit #`,
    rotten_weight = `Rotten fruitWeight (g)`,
    healthy_count = `Non-rotten fruit#`,
    healthy_weight = `Non-rotten fruitWeight (g)`,
    total_count = `Total #`,
    total_weight = `Total Weight (g)`
  ) %>%
  filter(!is.na(trt_string), !str_detect(trt_string, "A0")) %>%
  mutate(
    # "C" alone = treatment C (OTC); "AC","BC","CC","DC" = controls
    is_control = nchar(trt_string) >= 2 & str_detect(trt_string, "C$"),
    treatment = str_extract(trt_string, "^[A-D]"),
    plot_type = if_else(is_control, "Control", "OTC"),
    bio_rep = as.integer(bio_rep),
    pct_rotten_wt = rotten_weight / total_weight * 100,
    pct_rotten_ct = rotten_count / total_count * 100
  )

cat("Acute yield (excluding A0):", nrow(acute), "rows\n")
print(acute %>% count(cultivar, treatment, plot_type))
```

# 3 Exploratory Data Analysis

## 3.1 Long-term: OTC vs Control

```{r lt-eda, fig.height=4}
# Healthy weight by treatment and cultivar
ggplot(lt, aes(x = heat_treatment, y = healthy_weight, fill = heat_treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Long-term: Healthy Fruit Weight",
       x = NULL, y = "Weight (g)") +
  theme_minimal() + theme(legend.position = "none")
```

For Stevens, OTC and Control healthy weights are nearly identical (~684g vs ~681g).
For MQ, OTC yield is lower than Control (835g vs 961g), suggesting MQ may be
more sensitive to long-term heating, though variability is high with only 4 reps.

```{r lt-rotten, fig.height=4}
# Rotten percentage
ggplot(lt, aes(x = heat_treatment, y = pct_rotten_wt, fill = heat_treatment)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.1, size = 2) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Long-term: % Rotten by Weight",
       x = NULL, y = "% Rotten") +
  theme_minimal() + theme(legend.position = "none")
```

Interestingly, OTC plots have lower rotten rates in both cultivars. MQ has
higher rotten rates overall (2-8%) compared to Stevens (0-4%). The OTC
structure may provide some protection against rot-causing conditions.

## 3.2 Acute: Treatment Effect

```{r acute-hw, fig.height=5}
# Healthy weight by treatment, split by OTC/Control and cultivar
ggplot(acute, aes(x = treatment, y = healthy_weight, fill = plot_type)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(aes(color = plot_type), position = position_jitterdodge(jitter.width = 0.1),
              size = 1.5) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  scale_color_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Acute: Healthy Fruit Weight by Treatment",
       x = "Treatment (A=4, B=3, C=2, D=1 pulse)", y = "Weight (g)") +
  theme_minimal() + theme(legend.position = "bottom")
```

OTC (orange) consistently produces lower healthy weight than Control (blue)
across nearly all treatments and cultivars. The effect is visually clear.
However, there is no obvious dose-response pattern -- A (4 pulses) does not
yield notably less than D (1 pulse). St-B is the only case where OTC > Control.

```{r acute-rotten, fig.height=5}
# Rotten percentage
ggplot(acute, aes(x = treatment, y = pct_rotten_wt, fill = plot_type)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(aes(color = plot_type), position = position_jitterdodge(jitter.width = 0.1),
              size = 1.5) +
  facet_wrap(~ cultivar) +
  scale_fill_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  scale_color_manual(values = c("Control" = "#0072B2", "OTC" = "#D55E00")) +
  labs(title = "Acute: % Rotten by Weight",
       x = "Treatment (A=4, B=3, C=2, D=1 pulse)", y = "% Rotten") +
  theme_minimal() + theme(legend.position = "bottom")
```

OTC plots show dramatically higher rotten percentages than Control in every
treatment group. Control rotten rates cluster around 1-5%, while OTC rates
range from 5-37%. Treatment D (1 pulse) shows the highest variability and
some extreme outliers (>25% rotten), suggesting single late-season pulses
may be particularly damaging.

## 3.3 Acute: Paired OTC-Control Difference

```{r acute-diff, fig.height=5}
# Compute OTC - Control difference per treatment/cultivar/bio_rep pair
acute_paired <- acute %>%
  select(cultivar, treatment, bio_rep, plot_type, healthy_weight, pct_rotten_wt) %>%
  pivot_wider(names_from = plot_type,
              values_from = c(healthy_weight, pct_rotten_wt)) %>%
  mutate(diff_hw = healthy_weight_OTC - healthy_weight_Control,
         diff_rotten = pct_rotten_wt_OTC - pct_rotten_wt_Control)

ggplot(acute_paired, aes(x = treatment, y = diff_hw, color = cultivar)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_jitter(width = 0.15, size = 3, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.4,
               position = position_dodge(0.4)) +
  labs(title = "Acute: OTC - Control Healthy Weight Difference",
       subtitle = "Negative = heat reduced yield",
       x = "Treatment", y = "Difference (g)") +
  theme_minimal() + theme(legend.position = "bottom")
```

Nearly all paired differences are negative (below the dashed zero line),
confirming that OTC heating reduces yield. The mean reduction is about 155g.
The crossbars show group means -- all treatments show negative means except
St-B. There is no clear trend with pulse frequency.

# 4 Long-term Yield Modeling

## 4.1 Paired Design

Long-term experiment has 8 set pairs (OTC + Control). We use the set pairing
structure: plots 1-4 are OTC paired with plots 5-8 as Control (Stevens),
and plots 9-12 OTC paired with 13-16 Control (MQ).

```{r lt-pair}
# Assign set pair ID based on plot number
lt <- lt %>%
  mutate(set_pair = case_when(
    plot %in% c(1, 5)   ~ 1,
    plot %in% c(2, 6)   ~ 2,
    plot %in% c(3, 7)   ~ 3,
    plot %in% c(4, 8)   ~ 4,
    plot %in% c(9, 13)  ~ 5,
    plot %in% c(10, 14) ~ 6,
    plot %in% c(11, 15) ~ 7,
    plot %in% c(12, 16) ~ 8
  ))

# Compute paired differences
lt_wide <- lt %>%
  select(cultivar, set_pair, heat_treatment, healthy_weight, pct_rotten_wt) %>%
  pivot_wider(names_from = heat_treatment,
              values_from = c(healthy_weight, pct_rotten_wt)) %>%
  mutate(diff_hw = healthy_weight_OTC - healthy_weight_Control,
         diff_rotten = pct_rotten_wt_OTC - pct_rotten_wt_Control)

print(lt_wide %>% select(cultivar, set_pair, diff_hw, diff_rotten))
```

Paired differences vary widely: some pairs show OTC yielding more (+190g for
St pair 1), others show large reductions (-363g for MQ pair 8). This high
within-cultivar variability limits detection of a consistent OTC effect.

## 4.2 Paired t-test

```{r lt-ttest}
# Healthy weight: paired t-test (pooling both cultivars)
cat("=== Paired t-test: Healthy Weight (OTC - Control) ===\n")
t_hw <- t.test(lt_wide$diff_hw, mu = 0)
print(t_hw)

cat("\n=== Paired t-test: % Rotten (OTC - Control) ===\n")
t_rot <- t.test(lt_wide$diff_rotten, mu = 0)
print(t_rot)

# By cultivar
cat("\n=== By Cultivar: Healthy Weight ===\n")
for (cv in c("St", "MQ")) {
  sub <- lt_wide %>% filter(cultivar == cv)
  tt <- t.test(sub$diff_hw, mu = 0)
  cat(cv, ": mean diff =", round(mean(sub$diff_hw), 1),
      "g, p =", round(tt$p.value, 4), "\n")
}

cat("\n=== By Cultivar: % Rotten ===\n")
for (cv in c("St", "MQ")) {
  sub <- lt_wide %>% filter(cultivar == cv)
  tt <- t.test(sub$diff_rotten, mu = 0)
  cat(cv, ": mean diff =", round(mean(sub$diff_rotten), 1),
      "pp, p =", round(tt$p.value, 4), "\n")
}
```

None of the paired t-tests are significant (all p > 0.13). The overall mean
difference is -62g (OTC lower) but p=0.34. With only 4 pairs per cultivar,
statistical power is very limited. The direction suggests OTC may reduce MQ
yield (-126g) but has essentially no effect on Stevens (+3g).

## 4.3 Blocking ANOVA (set pair as block)

With only 8 set pairs, we treat set_pair as a fixed blocking factor to
account for the paired design. This is more appropriate than ignoring the
pairing structure.

```{r lt-block}
# Blocking ANOVA: set pair absorbs within-pair variability
m_lt_hw <- lm(healthy_weight ~ heat_treatment * cultivar + as.factor(set_pair),
              data = lt)
cat("=== Long-term: Healthy Weight (Blocking ANOVA) ===\n")
print(summary(m_lt_hw))

m_lt_rot <- lm(pct_rotten_wt ~ heat_treatment * cultivar + as.factor(set_pair),
               data = lt)
cat("\n=== Long-term: % Rotten (Blocking ANOVA) ===\n")
print(summary(m_lt_rot))
```

By including set_pair as a blocking factor, we absorb variability due to
spatial differences between plots. The OTC effect and cultivar effect are
estimated after controlling for this block-level variation.

## 4.4 Permutation Test (distribution-free validation)

With n=16, normality assumptions are hard to verify. We use permutation
tests as a non-parametric check.

```{r lt-perm}
# Permutation test for OTC effect on healthy weight
cat("=== Permutation test: Healthy Weight ~ OTC ===\n")
perm_hw <- independence_test(healthy_weight ~ factor(heat_treatment) | factor(cultivar),
                             data = lt)
print(perm_hw)

cat("\n=== Permutation test: % Rotten ~ OTC ===\n")
perm_rot <- independence_test(pct_rotten_wt ~ factor(heat_treatment) | factor(cultivar),
                              data = lt)
print(perm_rot)
```

The permutation tests provide distribution-free p-values, serving as a
robustness check for the parametric results above.

# 5 Acute Yield Modeling

## 5.1 Model: Healthy Weight

```{r acute-model-hw}
# Main effects + interaction
m_acute_hw <- lm(healthy_weight ~ treatment * cultivar + plot_type,
                 data = acute)
cat("=== Acute: Healthy Weight ===\n")
print(summary(m_acute_hw))
print(anova(m_acute_hw))
```

The ANOVA table shows that only plot_type (OTC vs Control) is significant
(p=0.001). Treatment frequency and cultivar are not significant, nor is the
treatment-by-cultivar interaction (p=0.73).

## 5.2 Model: % Rotten

```{r acute-model-rot}
m_acute_rot <- lm(pct_rotten_wt ~ treatment * cultivar + plot_type,
                  data = acute)
cat("=== Acute: % Rotten ===\n")
print(summary(m_acute_rot))
print(anova(m_acute_rot))
```

Same pattern: plot_type is dominant (p < 0.001). Treatment and cultivar
effects are not significant, and the interaction is weak (p=0.55).

## 5.3 Simplified Models (drop non-significant interaction)

```{r acute-simplified}
# Check if interaction is needed
cat("=== Interaction test: Healthy Weight ===\n")
m_hw_no_int <- lm(healthy_weight ~ treatment + cultivar + plot_type,
                  data = acute)
print(anova(m_hw_no_int, m_acute_hw))

cat("\n=== Interaction test: % Rotten ===\n")
m_rot_no_int <- lm(pct_rotten_wt ~ treatment + cultivar + plot_type,
                   data = acute)
print(anova(m_rot_no_int, m_acute_rot))
```

The F-tests confirm the interaction terms are not needed (p=0.73 and p=0.55).
We proceed with the additive models.

```{r acute-final}
# Final additive model for healthy weight
cat("=== Final Model: Healthy Weight ===\n")
print(summary(m_hw_no_int))
```

**Healthy weight model**: OTC reduces yield by 155g (p=0.001). Treatment
frequency has no significant effect -- B (+14g), C (+17g), D (+72g) relative
to A are all non-significant (p > 0.2). Cultivar is marginally non-significant
(St -69g, p=0.12). R-squared = 0.29.

## 5.4 Quasibinomial GLM for Rotten Rate

Rotten rate is proportion data (rotten count / total count). Instead of
transforming and using OLS, we model it directly with a Quasibinomial GLM,
which properly handles the bounded nature and variance structure of proportions.

```{r acute-glm-rot}
# GLM: model rotten count out of total count
m_glm_rot <- glm(cbind(rotten_count, healthy_count) ~
                   treatment + cultivar + plot_type,
                 family = quasibinomial, data = acute)
cat("=== Quasibinomial GLM: Rotten Rate ===\n")
print(summary(m_glm_rot))
```

```{r glm-interpret}
# Convert log-odds coefficients to odds ratios for interpretation
or_table <- data.frame(
  term = names(coef(m_glm_rot)),
  log_odds = round(coef(m_glm_rot), 3),
  odds_ratio = round(exp(coef(m_glm_rot)), 3),
  p_value = round(summary(m_glm_rot)$coefficients[, 4], 4)
)
cat("=== Odds Ratios ===\n")
print(or_table, row.names = FALSE)
```

The Quasibinomial GLM directly models the probability of a berry being
rotten. Coefficients are on the log-odds scale; odds ratios > 1 mean
higher rotten probability. This avoids the normality issues that plagued
the linear model on raw percentages.

# 6 Model Diagnostics

## 6.1 Long-term: Healthy Weight

```{r diag-lt, fig.height=7}
par(mfrow = c(2, 2))
plot(m_lt_hw, which = 1:4)
```

Residuals vs Fitted shows no obvious pattern. QQ plot looks acceptable for
n=16. No points exceed Cook's distance threshold. The diagnostics are
adequate given the small sample size.

## 6.2 Acute: Healthy Weight

```{r diag-acute-hw, fig.height=7}
par(mfrow = c(2, 2))
plot(m_hw_no_int, which = 1:4)
```

Residuals are approximately normal (QQ plot follows the line well). Residuals
vs Fitted shows relatively constant variance. A few points (rows 22, 24, 37)
have moderate leverage but Cook's distances remain below the threshold. This
model satisfies linear regression assumptions reasonably well.

## 6.3 Acute: Quasibinomial GLM Diagnostics

```{r diag-glm-rot, fig.height=7}
par(mfrow = c(2, 2))
plot(m_glm_rot, which = 1:4)
```

The GLM diagnostics look reasonable. Residuals vs Fitted shows no strong
pattern. The QQ plot is much better than the linear model on raw percentages
because the quasibinomial family naturally accounts for the skewed, bounded
nature of proportion data. A few high-leverage points remain (the same plots
with extreme rotten rates) but Cook's distances are acceptable.

## 6.4 Influence Check

```{r influence}
# Check high Cook's distance points in GLM
cooks_glm <- cooks.distance(m_glm_rot)
influential <- which(cooks_glm > 4 / nrow(acute))
cat("Influential points (Cook's D > 4/n):\n")
print(acute[influential, c("cultivar", "trt_string", "bio_rep",
                            "healthy_weight", "pct_rotten_wt")])
```

These are real observations, not errors -- they represent plots where acute
heat caused severe fruit damage. We retain them in the analysis.

# 7 Summary of Results

## 7.1 Long-term Experiment

```{r summary-lt, echo=FALSE}
cat("Long-term Heat Experiment (n=16)\n")
cat("Models: Blocking ANOVA + Paired t-test + Permutation test\n\n")

cat("Key findings:\n")
cat("- Cultivar is the dominant factor: St produces ~280g less",
    "than MQ (p=0.002)\n")
cat("- OTC effect on yield: not significant in any model\n")
cat("  MQ: -126g (p=0.11), St: +3g (p=0.97)\n")
cat("- OTC reduced rotten % marginally (p=0.064)\n")
cat("- Permutation tests confirm the parametric results\n")
cat("- Limited power with n=4 per group\n")
```

## 7.2 Acute Experiment

```{r summary-acute, echo=FALSE}
cat("Acute Heat Experiment (n=48, 3 bio reps per group)\n")
cat("Models: ANOVA for healthy weight, Quasibinomial GLM for",
    "rotten rate\n\n")

cat("Key findings:\n")
cat("1. OTC heating significantly REDUCED healthy fruit weight",
    "by 155g (p=0.001)\n")
cat("   Baseline (MQ, Treatment A, Control): ~1011g\n")
cat("   OTC effect: -155g (~15% reduction)\n\n")

cat("2. OTC heating significantly INCREASED rotten probability\n")
cat("   Quasibinomial GLM: OTC odds ratio for rotten fruit\n")
cat("   interpreted on log-odds scale (see Section 5.4)\n\n")

cat("3. Treatment frequency (A=4, B=3, C=2, D=1 pulses) did",
    "NOT significantly affect yield\n")
cat("   No dose-response relationship detected\n\n")

cat("4. Cultivar differences: Stevens tends to have lower yield",
    "and lower rotten rate\n\n")

cat("Limitations:\n")
cat("- Small sample size (n=3 per treatment-cultivar-plot_type)\n")
cat("- A few influential observations with very high rotten rates",
    "(27-37%)\n")
```
